---
title: "The {xdvir} Package"
author: "Paul Murrell"
output: 
  rmarkdown::html_vignette:
    number_sections: true
bibliography: xdvir-intro.bib
vignette: >
  %\VignetteIndexEntry{The {xdvir} Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = ''
)
```

*The {xdvir} package provides functions for rendering LaTeX snippets in
R plots.*

```{r}
library(xdvir)
```

# Rendering LaTeX in R

The plot below, produced with R, features a LaTeX-quality
mathematical equation in the
top-left corner.
Producing an equation like this requires three things:

1.  A **markup** language that allows us to describe the equation. 
    This is complicated by the fact
    that the equation is not simply one character placed after another from
    left to right along a line and the fact that
    elements of the equation are not simple characters.
    For example, we need to be able to express that the digit "2"
    needs to be placed as a superscript relative to the character "z"
    and we need to be able to express the greek letter pi.

2.  A **typesetter** that can select fonts and individual 
     glyphs from within those fonts and determine the exact placement
     of each individual glyph.

3.  A **renderer** that can draw the glyphs that the typesetter
    has selected in the locations that the typesetter has determined.

```{r, label="tex", echo=FALSE}
tex <- r"(\huge $\Phi(z) = \frac{1}{\sqrt{2\pi}} \cdot e^{-\frac{z^2}{2}}$)"
```
```{r, echo=FALSE, fig.width=6, fig.height=3, message=FALSE}
library(gggrid)
x <- seq(-4, 4, length.out=100)
df <- data.frame(x=x, y=dnorm(x))
ggplot(df) +
    geom_area(aes(x, y), colour=NA, fill="grey90") +
    geom_line(aes(x, y)) +
    grid_panel(latexGrob(tex, x=0, y=1, hjust="left", vjust="top")) +
    xlim(-4, 4) +
    scale_x_continuous(expand=expansion(0)) +
    scale_y_continuous(expand=expansion(c(0))) +
    coord_cartesian(clip="off") +
    theme_minimal() +
    theme(panel.grid=element_blank(),
          axis.title=element_blank(),
          axis.text.y=element_blank(),
          plot.margin=unit(rep(4, 4), "mm")) 
```

The {xdvir} package allows us to use LaTeX as the **markup** language,
a TeX engine as the **typesetter**, and R as the **renderer**.
This allows us to produce high-quality LaTeX typesetting
within R plots, as in the image above.

The main function in {xdvir} is `grid.latex()`,
which takes a LaTeX snippet and draws it on the 
current R graphics device.
For example, the following code
describes the mathematical equation, using LaTeX code, as a character value,
then draws it by calling `grid.latex()` with the character value
as the first argument.

Because LaTeX code tends to contain 
a large number of backslashes, the code below uses the `r"(...)"`
syntax for raw character constants, so that we do not have to
escape each backslash with a double backslash.

```{r}
<<tex>>
```

```{r fig.width=3, fig.height=1}
grid.latex(tex)
```

# Rendering LaTeX in plots

As the function name `grid.latex()`
suggests, {xdvir} only provides a low-level
interface for drawing LaTeX snippets as part of {grid} output.
However, there are ways to combine {grid} output with higher-level
plotting packages.

For example, we can use {grid} within panel functions in the
{lattice} package [@pkg-lattice].  The following code draws a {lattice} plot
and adds the mathematical equation in the top-left corner.
This demonstrates that the `grid.latex()` function has
arguments `x`, `y`, `hjust`, and `vjust` for positioning the
LaTeX output.

```{r}
library(lattice)
```

```{r, label="lattice-setup", echo=FALSE, eval=FALSE}
lattice.options(layout.widths=list(left.padding=list(x=0),
                                   right.padding=list(x=0)),
                layout.heights=list(bottom.padding=list(x=0),
                                    top.padding=list(x=0)))
```

```{r, label="lattice-plot", eval=FALSE}
xyplot(y ~ x, df, type="l",
       panel=function(...) {
           panel.xyplot(...)
           grid.latex(tex, 
                      x=unit(2, "mm"),
                      y=unit(1, "npc") - unit(2, "mm"),
                      hjust="left", vjust="top")
       })
```

```{r, echo=FALSE, fig.width=6, fig.height=3}
<<lattice-setup>>
<<lattice-plot>>
```

If we want to add {grid} to a {graphics} plot, we can use the
{gridGraphics} package [@RJ-2015-012; @pkg-gridGraphics].
The following code draws a {graphics} plot, uses `gridGraphics::grid.echo()` to
convert it to {grid}, navigates to the {grid} viewport that
represents the main plot region, and draws the mathematical equation
in the top-left corner.

```{r}
library(gridGraphics)
```

```{r, label="graphics-setup", echo=FALSE, eval=FALSE}
par(mar=c(4, 4, 1, 1))
```
```{r, label="graphics-plot", eval=FALSE}
plot(y ~ x, df, type="l")
grid.echo()
downViewport("graphics-plot-1")
grid.latex(tex, 
           x=unit(2, "mm"),
           y=unit(1, "npc") - unit(2, "mm"),
           hjust="left", vjust="top")
```
```{r, echo=FALSE, fig.width=6, fig.height=3, fig.keep="last"}
<<graphics-setup>>
<<graphics-plot>>
```

We can add {grid} output to {ggplot2} plots [@pkg-ggplot2]
with the {gggrid}
package [@murrell-gggrid-2021; @pkg-gggrid].
The following
code uses the `latexGrob()` function to generate a "graphical object"
for {ggplot2} to draw, rather than drawing the output immediately as
`grid.latex()` does, and passes that to `gggrid::grid_panel()`
so that the mathematical equation is added in the top-left corner
of the plot.  


```{r, message=FALSE}
library(gggrid)
```

```{r, fig.width=6, fig.height=3}
ggplot(df) +
    geom_line(aes(x, y)) +
    grid_panel(latexGrob(tex,
                         x=unit(2, "mm"), 
                         y=unit(1, "npc") - unit(2, "mm"), 
                         hjust="left", vjust="top"))
```

# Limitations

Need device with glyph support

# Trouble shooting

Detecting TeX (and installing TeX)

# References
